Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-05-01 paper/paper.tex
a major pass on RL sections

In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
\textit{Markov decision processes (MDPs)}\citep{sutton1998reinforcement}.
For the remainder of this paper, for the sake of simplicity
we use the abbreviation ``MDP'' to refer to the special case
of \textit{Partially Observable Markov Decision Processes}.
A MDP operationalizes a sequential decision process
in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., is likely to be different for another agent in the same state).
This is performed alongside each transition in the environments state space, and maintains a probability distribution over the set of possible
states, based on a set of observations and observation probabilities. This is a minimal amount of assumptions that can be made about an environment,
and is characteristic of so-called \textit{model-free RL},
which we propose as a useful formal framework for DMN function.
%

\paragraph{Model-free reinforcement learning.}
Model-free RL has had great success in real-world problems including robotics \citep{abbeel2004,andrewng}, computer game play, \citep{mnih2015,silver2016mastering},
etc., just to name a few \textbf{XXX Please expand with 1-2 additional sentences}.
Such a learning scheme can be
plausibly realized in the human brain \citep{doherty2015structure}. Indeed,
a core property of human intelligence has been proposed to underly improvement
of expected utility outcomes as a strategy for action choice in uncertain
environments \citep{gershman2015computational}, a situation perfectly captured by the formalism of MDPs.
It has also long been proposed \citep{dayan2008decision} that there
is a rather direct mapping of model-free RL learning algorithms
onto aspects of the brain.
The neurotransmitter dopamine could serve
as a ``teaching signal'' to better estimate value associations
and action policies by controlling
synaptic plasticity in the reward system, including the NAc.
In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's environment or constraints on the state space and transitions between states.
% For instance, if a rat in a maze knows
% that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
%

@@

\begin{definition*}
Mathematically, a MDP is simply a quintuplet $(\mathcal S, \mathcal A, r, p, \mu)$ where
\begin{itemize}
\item $\mathcal S$ is the set of states, such as $\mathcal S = \{\text{happy}, \text{sad}\}$.
\item $p : \mathcal S \times \mathcal A \times \mathcal S \rightarrow [0, 1],\; (s,a,s') \mapsto p(s'|s,a)$,
  the probability of moving to state $s'$ if action $a$ is taken from state $s$. In addition, one requires that such
  transitions be Markovian. Consequently, the future states are independent of past states and only depend on the present.
\item $\mathcal A$ is the set of actions, such as $\mathcal A = \{\text{read}, \text{run},
  \text{laugh}, \text{sympathize}, \text{empathize}\}.$
\item $r : \mathcal S \times \mathcal A \rightarrow \mathbb R$ is the \textit{reward function},
  so that $r(s, a)$ is the instant reward for taking action $a$ in state $s$.
\item $\mu: \mathcal S \rightarrow [0, 1]$ is the prior probability on the states so that
  $\mu(s)$ is the probability that the environment starts off in state $s$.
\end{itemize}

@@

\subsection{The nuts and bolts of reinforcement learning in the DMN}
\textbf{XXX Elvis to Elvis a short preface is needed here, explaining the roadmap of this subsection from a high-level perspective!}

\subsubsection{The objective value of an action given a state}
Given only the limited information available from a MDP, at a state $s$ the average
utility of choosing an action $a$ under a policy $\pi$ can be captured by the single number
% We can now formulate the purpose of the DMN
% as finding a policy $\pi$ for the human agent that maximizes the
% expected cumulative value of an state-action pair $(a,s)$, also known as the \textit{Q-value}, namely
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s,a) = \mathbb E [G^\pi|s_0=s,a_0=a],
\end{equation}
called the $Q$-value for the state-action pair $(s,a)$.
In other words, $Q^{\pi}(s,a)$ corresponds to the expected reward
over all considered action trajectories, in which
the agent sets out in the environment in state
$s$, chooses action $a$, and then follows the policy $\pi$ to select future actions.
%% Given that each $\pi$ correspondence to an exhaustive description of
%% behavioral patterns, $Q^{\pi}(s_t,a_t)$ corresponds to the epistemic value of
%% the set of actions given states in finite time.
For the brain,
$Q^{\pi}(s, a)$ defined in \eqref{eq:qval} provides the subjective
utility of executing a specific action.
Thus, it answers the question
``What is the expected utility of taking action $a$ in this situation ?''.
$Q^{\pi}(s,a)$ offers a formalization of optimal behavior that
may well capture processing aspects of the DMN in human agents.

@@

\begin{equation}
  \pi^{\text{greedy}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
by querying an oracle \textbf{XXX Please let's not use "oracle" anywhere in
the paper, if possible, because it will confuse a lot of peoople.}
which maps from states directly to actions. Here $\tilde{Q}(s,a|\theta)$ is an
approximate version of the $Q$-value function parameterized by the tensor $\theta$.
% $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$.
For instance, a simplistic linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as proposed by \citep{songNIPS2016}.
%  and Parr et al. 2003 (JMLR)].
In the DMN, the dmPFC would implement such a greedy lookup
over the action space.
The model
  parameters $\theta$ would correspond to synaptic weights and connection strengths between
  brain regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the greedy policy \eqref{eq:policy}.
  The oracle Q-values $\tilde{Q}(s, a|\theta)$ tell the DMN the (expected) usefulness of taking an action a in state $s$.
  The DMN, and in particular its dmPFC node, could then contribute to the choice, at a given state $s$, of an action $a$ which maximizes this tabulated values. This mapping from states to actions is conventionally called ``policy matrix''.
  Learning consists in starting from a given table and
  updating it during action choices,
  which take the agent to different table entries.

@@

\subsubsection{Optimal control via stochastic gradient descent in the DMN}
Efficient learning of the entire set of model parameters can effectively be achieved
via \textit{gradient descent}.
\textbf{XXX Please add one explanatory sentence for GD.}
Reward prediction errors \textemdash also known as \textit{regret}\textemdash percolate
from lower to higher processing layers to modulate the choice of future actions

@@

% \paragraph{Parameter update.}
\begin{equation}
  \delta \theta_{k} \propto -\nabla_{\theta_{k}}\mathcal L(\theta_{k})
  = -\mathbb E_{(s, a, r, s') \sim \mathcal D}[\underbrace{(\tilde{Q}(s, a|\theta_{k}) - y_k)}_{\text{prediction error}}
    \underbrace{\nabla_{\theta_{k}}\tilde{Q}(s, a|\theta_{k})}_{\text{aversion}}].
  \label{eq:oracle}
\end{equation}
Over time (i.e., in the limit $k \rightarrow \infty$),
the thus generated approximating sequence of
$Q$-value functions provably
% \footnote{The proof is standard and relies on the Banach Fixed-Point Theorem applied to the Bellman transform.}
$\tilde{Q}(.,.|\theta_0) \rightarrow \tilde{Q}(.,.|\theta_1) \rightarrow \tilde{Q}(.,.|\theta_2) \rightarrow \ldots$
are attracted and absorbed by the optimal $Q$-value function $Q^*$ defined implicitly by the Bellman equation \eqref{eq:bellman}.
\textbf{XXX Please say something about *stochasti*, which you can then
also use as transition into the next passage.}

\subsubsection{Link to classical reinforcement learning algorithms}
One should note that most classical RL algorithms, including \textit{Temporal Difference (TD)}
learning \citep{sutton1998reinforcement}, REINFORCE \citep{williams1992}, and SARSA can be
cast in this \textbf{please unpack the "this" here}
general variational framework. For instance, TD corresponds to
the above framework using a linear value approximator with feature encoding
$\phi(s,a) = \delta_{(s,a)} =  \text{ point mass at }(s,a)$ on the grid
$\mathcal S \times \mathcal A$, and so
$$\nabla_{\theta}\tilde{Q}(s, a|\theta) = \phi(s, a) = \delta_{(s,a)}.$$
Whence, the gradient update due to the sample $(s,a,r,s') \in \mathcal D$ is
$$\tilde{Q}(s, a) \leftarrow \tilde{Q}(s, a) + \alpha(y-\tilde{Q}(s, a)),$$
the well-known TD update rule \citep{sutton1998reinforcement}.

@@

\section{More on RL in the brain}
\subsection{Does hippocampal replay equal inverse reinforcement-learning?}
%\begin{mdframed}
  Given the trace $s_0,a_0,s_1,\ldots,$ of an optimal agent's strategy $\pi^*$ in
  a MDP (called a \textit{teacher's demonstration}), can we figure out what is the
  (instantaneous) reward function $r: \mathcal S \times \mathcal \mathcal A
  \rightarrow \mathbb R$ that the agent is optimizing over a prescribed class of reward functions
  (e.g., linear rewards $r(s,a) \equiv \theta^T\phi(s,a)$)?
  For instance, given traces of motor actions from an adult to grab a cup on a table,
  can an observing child figure out what ``langragian'' functional is being minimized by the former?
  How? Can they reproduce this optimal behavior?
%\end{mdframed}
Such questions are of course pertinent to our decision-making theory for the DMN.
In the general artificial case, the problem has been extensively studied and partially solved by
\citep{abbeel2004}. They are also been rigorously studied in general optimal control literature under the name
``inverse optimal control'', but in model-based certain (where the physical dynamics are known, etc.)...

>>>>

In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
\textit{Markov decision processes (MDPs)}\citep{sutton1998reinforcement}.
% For the remainder of this paper, for the sake of simplicity
% we use the abbreviation ``MDP'' to refer to the special case
% of \textit{Partially Observable Markov Decision Processes}.
A MDP operationalizes a sequential decision process
in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., is likely to be different for another agent in the same state).
This is performed alongside each transition in the environments state space, and maintains a probability distribution over the set of possible
states, based on a set of observations and observation probabilities. This is a minimal amount of assumptions that can be made about an environment,
and is characteristic of so-called \textit{model-free RL}. Model-free RL has had great success in real-world problems including robotics
\citep{abbeel2004,andrewng}, computer game play, \citep{mnih2015,silver2016mastering}, etc., just to name a few.
We propose to use such a model as formal basis to model DMN function.

Such a learning scheme can be plausibly realized in the human brain \citep{doherty2015structure}. Indeed,
a core property of human intelligence has been proposed to underly improvement
of expected utility outcomes as a strategy for action choice in uncertain
environments \citep{gershman2015computational}, a situation perfectly captured by the formalism of MDPs.
It has also long been proposed \citep{dayan2008decision} that there
is a rather direct mapping of model-free RL learning algorithms
onto aspects of the brain.
The neurotransmitter dopamine could serve
as a ``teaching signal'' to better estimate value associations
and action policies by controlling
synaptic plasticity in the reward system, including the NAc.
In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's environment or constraints on the state space and transitions between states.
% For instance, if a rat in a maze knows
% that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
%

@@

\begin{definition*}
Mathematically, an MDP is simply a quintuplet $(\mathcal S, \mathcal A, r, p, \mu)$ where
\begin{itemize}
\item $\mathcal S$ is the set of states, such as $\mathcal S = \{\text{happy}, \text{sad}\}$.
\item $p : \mathcal S \times \mathcal A \times \mathcal S \rightarrow [0, 1],\; (s,a,s') \mapsto p(s'|s,a)$,
  the probability of moving to state $s'$ if action $a$ is taken from state $s$. In addition, one requires that such
  transitions be Markovian. Consequently, the future states are independent of past states and only depend on the present.
\item $\mathcal A$ is the set of actions, such as $\mathcal A = \{\text{read}, \text{run},
  \text{laugh}, \text{sympathize}, \text{empathize}\}.$
\item $r : \mathcal S \times \mathcal A \rightarrow \mathbb R$ is the \textit{reward function},
  so that $r(s, a)$ is the instant reward for taking action $a$ in state $s$.
\item $\mu: \mathcal S \rightarrow [0, 1]$ is the prior probability on the states so that
  $\mu(s)$ is the probability that the environment starts off in state $s$.
\end{itemize}

@@

\subsection{The nuts and bolts of reinforcement learning in the DMN}
Given only the limited information available from an MDP, at a state $s$ the average
utility of choosing an action $a$ under a policy $\pi$ can be captured by the single number
% We can now formulate the purpose of the DMN
% as finding a policy $\pi$ for the human agent that maximizes the
% expected cumulative value of an state-action pair $(a,s)$, also known as the \textit{Q-value}, namely
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s,a) = \mathbb E [G^\pi|s_0=s,a_0=a],
\end{equation}
called the $Q$-value for the state-action pair $(s,a)$.
In other words, $Q^{\pi}(s,a)$ corresponds to the expected reward
over all considered action trajectories, in which
the agent sets out in the environment in state
$s$, chooses action $a$, and then follows the policy $\pi$ to select future actions.
%% Given that each $\pi$ correspondence to an exhaustive description of
%% behavioral patterns, $Q^{\pi}(s_t,a_t)$ corresponds to the epistemic value of
%% the set of actions given states in finite time.
For the brain,
$Q^{\pi}(s, a)$ defined in \eqref{eq:qval} provides the subjective
utility of executing a specific action.
Thus, it answers the question
``What is the expected utility of taking action $a$ in this situation ?''.
$Q^{\pi}(s,a)$ offers a formalization of optimal behavior that
may well capture processing aspects of the DMN in human agents.

@@

\begin{equation}
  \pi^{\text{greedy}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
by querying an an approximate version $\tilde{Q}(s,a|\theta)$ of the $Q$-value function, parameterized
by the tensor $\theta$. This defines a mapping from states directly to actions.
% $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$.
For instance, a simplistic linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as proposed by \citep{songNIPS2016}.
%  and Parr et al. 2003 (JMLR)].
In the DMN, the dmPFC would implement such a greedy lookup
over the action space.
The model
  parameters $\theta$ would correspond to synaptic weights and connection strengths between
  brain regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the greedy policy \eqref{eq:policy}.
  The oracle Q-values $\tilde{Q}(s, a|\theta)$ tell the DMN the (expected) usefulness of taking an action a in state $s$.
  The DMN, and in particular its dmPFC node, could then contribute to the choice, at a given state $s$, of an action $a$ which maximizes this tabulated values. This mapping from states to actions is conventionally called ``policy matrix''.
  Learning consists in starting from a given table and
  updating it during action choices,
  which take the agent to different table entries.

@@

\subsubsection{Optimal control via stochastic gradient descent in the DMN}
Efficient learning of the entire set of model parameters can effectively be achieved
via stochastic \textit{gradient descent}, a first-order\footnote{ i.e using only differential information up to order 1.} algorithm for finding
local minimal of differentiable functions. Stochastic here that the true gradient is estimated from batches of training examples samples from the replay memory, the HC

@@

% \paragraph{Parameter update.}
\begin{equation}
  \delta \theta_{k} \propto -\nabla_{\theta_{k}}\mathcal L(\theta_{k})
  = -\mathbb E_{(s, a, r, s') \sim \mathcal D}[\underbrace{(\tilde{Q}(s, a|\theta_{k}) - y_k)}_{\text{prediction error}}
    \underbrace{\nabla_{\theta_{k}}\tilde{Q}(s, a|\theta_{k})}_{\text{aversion}}].
  \label{eq:oracle}
\end{equation}
Thus actions are taken to drive reward prediction errors \textemdash also known as \textit{regret}\textemdash percolate
from lower to higher processing layers to modulate the choice of future actions .Over time (i.e., in the limit $k \rightarrow \infty$),
the thus generated approximating sequence of
$Q$-value functions provably
% \footnote{The proof is standard and relies on the Banach Fixed-Point Theorem applied to the Bellman transform.}
$\tilde{Q}(.,.|\theta_0) \rightarrow \tilde{Q}(.,.|\theta_1) \rightarrow \tilde{Q}(.,.|\theta_2) \rightarrow \ldots$
are attracted and absorbed by the optimal $Q$-value function $Q^*$ defined implicitly by the Bellman equation \eqref{eq:bellman}.

One should note that most classical RL algorithms, including \textit{Temporal Difference (TD)}
learning \citep{sutton1998reinforcement}, REINFORCE \citep{williams1992}, and SARSA can be
cast in the general variational framework \eqref{eq:oracle}. For instance, TD corresponds to
the above framework using a linear value approximator with feature encoding
$\phi(s,a) = \delta_{(s,a)} =  \text{ point mass at }(s,a)$ on the grid
$\mathcal S \times \mathcal A$, and so
$$\nabla_{\theta}\tilde{Q}(s, a|\theta) = \phi(s, a) = \delta_{(s,a)}.$$
Whence, the gradient update due to the sample $(s,a,r,s') \in \mathcal D$ is
$$\tilde{Q}(s, a) \leftarrow \tilde{Q}(s, a) + \alpha(y-\tilde{Q}(s, a)),$$
the well-known TD update rule \citep{sutton1998reinforcement}.

@@

\section{More on RL in the brain}
\subsection{Does hippocampal replay equal inverse reinforcement-learning?}
%\begin{mdframed}
  Given the trace $s_0,a_0,s_1,\ldots,$ of an optimal agent's strategy $\pi^*$ in
  an MDP (called a \textit{teacher's demonstration}), can we figure out what is the
  (instantaneous) reward function $r: \mathcal S \times \mathcal \mathcal A
  \rightarrow \mathbb R$ that the agent is optimizing over a prescribed class of reward functions
  (e.g., linear rewards $r(s,a) \equiv \theta^T\phi(s,a)$)?
  For instance, given traces of motor actions from an adult to grab a cup on a table,
  can an observing child figure out what ``langragian'' functional is being minimized by the former?
  How? Can they reproduce this optimal behavior?
%\end{mdframed}
Such questions are of course pertinent to our decision-making theory for the DMN.
In the general artificial case, the problem has been extensively studied and partially solved by
\citep{abbeel2004}. They are also been rigorously studied in general optimal control literature under the name
``inverse optimal control'', but in model-based certain (where the physical dynamics are known, etc.)...
